%==============================================================================
% System-Heterogeneity section — drop-in LaTeX for Overleaf
% Sibling to thesis_ready/writing/09_overleaf_ready.tex (statistical het).
% Copy this entire file as a separate .tex document in the same Overleaf
% project, OR \input it into your main thesis document after the statistical
% heterogeneity section.
%
% All HPC-pending values use \TODOhpcSH{} markers. Toggle \showtodoshtrue/false
% near the top to render markers visibly (default) or hide them.
%
% Compiles standalone (with \documentclass etc. at top) or as an \input{...}
% chunk (delete the document scaffolding).
%==============================================================================
\documentclass[11pt,a4paper]{article}

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{amsmath, amssymb}
\usepackage{microtype}
\usepackage{hyperref}
\usepackage{xcolor}
\usepackage[margin=2.5cm]{geometry}
\usepackage{setspace}
\onehalfspacing

\providecommand{\citep}[1]{(#1)}
\providecommand{\citet}[1]{#1}

% --- TODO markers for HPC-pending system-het values -------------------------
% These are deliberately distinct from \TODOhpc in the statistical-het file,
% so the two sections can co-exist in the same Overleaf project without
% macro collisions if you choose to merge them.
\newif\ifshowtodosh
\showtodoshtrue
\newcommand{\TODOhpcSH}[1]{%
  \ifshowtodosh\textcolor{red}{\textbf{[SH-HPC: #1]}}\else#1\fi}

\title{System Heterogeneity: FedProx under Variable Client Compute Budgets}
\author{}
\date{}

\begin{document}
\maketitle

%==============================================================================
\section{Motivation and Background}
%==============================================================================

The statistical-heterogeneity experiment of the preceding section
demonstrated that FedProx \citep{Li et al., 2020} outperforms FedAvg
\citep{McMahan et al., 2017} on a deliberately non-IID partition of
DermaMNIST. That experiment isolated the data-distribution axis of
federated heterogeneity. The present section isolates the second canonical
axis identified by the original FedProx paper: \textbf{system
heterogeneity} -- variability in the local computational budget each
client can contribute per communication round.

System heterogeneity is, in fact, the \emph{primary} motivation for
FedProx's proximal regulariser. \citet{Li et al., 2020} proved (their
Theorem~4) a convergence guarantee for FedProx under $\gamma$-inexact
local updates: each client may submit a parameter vector $w_i^{t+1}$ that
does not fully minimise its local objective, provided the residual
inexactness is bounded. The proximal term $\frac{\mu}{2}\lVert w - w^t \rVert^2$
constrains how far an inexact local update can drift from the global
anchor $w^t$, making partial work from stragglers usable for aggregation
without destabilising the global model. FedAvg has no such guarantee, and
its size-weighted aggregation rule treats partial-work and full-work
updates symmetrically -- which biases the global model toward whichever
clients drifted furthest, regardless of how much work they actually
completed.

\subsection*{Why this complements the statistical-heterogeneity result}

The preceding section answered: \emph{does FedProx help when clients have
heterogeneous data distributions?} This section answers a logically
distinct question: \emph{does FedProx help when clients have heterogeneous
computational budgets?} Both are positive answers in the original Li et
al.\ paper, but on different benchmarks (Synthetic, FEMNIST, Shakespeare,
Sentiment140) -- not on medical imaging. Demonstrating both effects on
DermaMNIST under the same partition, same paired-seed protocol, and same
statistical machinery gives a coherent two-axis defence of FedProx's
theoretical motivation.

\subsection*{Scope: FedProx vs.\ FedNova for heterogeneous local steps}

This section restricts itself to FedAvg vs.\ FedProx, matching the
project brief and the protocol of the preceding statistical-heterogeneity
section. We acknowledge that for the specific case of \emph{heterogeneous
numbers of local update steps}, the more specialised competitor algorithm
is \textbf{FedNova} \citep{Wang et al., 2020}, which argues that
heterogeneous local steps create objective inconsistency under naïve
weighted aggregation and proposes a normalised-averaging rule to address
it. FedNova's mechanism is therefore directly engineered for the regime
we test in this section, while FedProx's proximal-anchor mechanism is
more general (it addresses both client-distribution heterogeneity and
straggler inexactness). A direct FedNova comparison is identified as
the most natural extension of this section and is recommended for
follow-up work; we do not implement it here.

%==============================================================================
\section{Experimental Setting}
%==============================================================================

\subsection{Design choices held identical to the statistical-heterogeneity sweep}

To isolate the system-heterogeneity contribution, every other experimental
parameter is held identical to the headline statistical-heterogeneity
sweep of the preceding section: same \texttt{balanced\_paired\_7\_clients}
partition, same 10 paired seeds $\{42, 123, 456, 789, 999, 2024,$
$31337, 8675309, 161803, 271828\}$, same \texttt{DermMNISTCNN}
architecture, same hyperparameters ($R = 150$, $\mu = 0.01$, batch size
32, SGD with $\eta = 0.01$, momentum $0.9$), and same evaluation
protocol (test set evaluated once per run at the best-val checkpoint).
The only variable is the per-client per-round local-epoch budget.

\subsection{Three conditions}

\paragraph{C0 (baseline, no system heterogeneity).} Every client performs
$E = 20$ local epochs every round. This is identical to the headline
statistical-heterogeneity sweep; its results are reported in the preceding
section ($M_\text{FedAvg} = 0.481 \pm 0.025$, $M_\text{FedProx} = 0.508 \pm 0.014$,
$\Delta = +0.027$, paired Wilcoxon $p = 0.020$, rank-biserial $r_{rb} = +0.818$).

\paragraph{C1 (fixed stragglers --- a specialist + a generalist).} Two
designated clients (C5 and C6 in the partition) always perform
$E_\text{straggler} = 5$ local epochs; the remaining five clients perform
$E = 20$. This models a deployment where specific institutions
consistently have weaker compute infrastructure. The 5/20 ratio mirrors
the configuration in \citet{Marija, 2025} §3.8.4 but adapted to a
7-client partition.

\emph{Important caveat on the C1 design.} C5 belongs to the
melanoma/vascular specialty pair (the per-specialty pair that drives the
headline FedProx advantage; see preceding section), and C6 is the
nevi-only generalist. C1 therefore conflates ``system heterogeneity'' with
``a particular clinical role configuration'' --- specifically, the
combination of a high-Δ specialist client and the generalist client both
running slow. Any C1 result must be interpreted as \emph{this particular
configuration} rather than a clean isolation of the system-heterogeneity
effect. The cleanly-isolating test is C2 below, which is also the
canonical Li et al.\ (2020) setup. C1 is retained because of its
operational realism (in real deployments, certain client identities
have predictable hardware constraints), but the H2 (between-condition)
inference in this section relies primarily on C2.

\paragraph{C2 (random stragglers, Li et al.\ 2020 protocol).} Each
communication round, half of the participating clients are randomly
designated stragglers. Each straggler performs $E_i \sim
\text{Uniform}[1, 19]$ local epochs; non-stragglers perform $E = 20$.
The straggler schedule is seeded with the same seed as the rest of the
run, so paired FedAvg/FedProx pairs share an identical schedule and the
within-pair $\Delta$ reflects only the algorithmic difference. This
matches the canonical FedProx experimental setup in \citet{Li et al., 2020} §5.2.

\subsection{Pre-registered hypotheses}

\paragraph{H1 (within-condition).} For each system-heterogeneity
condition $c \in \{C1, C2\}$, FedProx outperforms FedAvg on test macro-F1
under paired Wilcoxon ($\alpha = 0.05$, two-sided, $n = 10$).

\paragraph{H2 (between-condition).} The within-pair $\Delta$ is larger
under system heterogeneity than at baseline:
$\mathbb{E}[\Delta_s^{C1}] > \mathbb{E}[\Delta_s^{C0}]$ and
$\mathbb{E}[\Delta_s^{C2}] > \mathbb{E}[\Delta_s^{C0}]$. This is tested
by paired Wilcoxon on the per-seed difference $\Delta_s^c - \Delta_s^{C0}$
for each $c \in \{C1, C2\}$. A positive significant result supports the
canonical FedProx claim that the proximal regulariser is most beneficial
under inexact local updates. Because H2 comprises two related tests (one
per condition), we report both raw and Bonferroni-corrected $p$-values
($\alpha_\text{family} = 0.05$, corrected per-test threshold
$\alpha / 2 = 0.025$); the canonical interpretation rests on C2 (the Li et
al.\ 2020 setup) and is robust to whether C1 is also significant.

%==============================================================================
\section{Results}
%==============================================================================

\subsection{Headline test performance under system heterogeneity}

\begin{table}[h]
\centering
\caption{Test macro-F1 under three conditions of system heterogeneity,
10 paired seeds each. C0 (baseline) is reproduced from the preceding
section; C1 and C2 are pending HPC sweep completion. The within-condition
$\Delta$ tests H1; the between-condition $\Delta - \Delta_{C0}$ tests H2.}
\label{tab:sh-headline}
\small
\begin{tabular}{lccccc}
\toprule
Condition & FedAvg & FedProx & $\Delta$ & Wilcoxon $p$ (H1) & H2 test $p$ \\
\midrule
C0 (no system het, baseline)   & $0.481 \pm 0.025$ & $0.508 \pm 0.014$ & $+0.027$ & $0.020$ & (reference) \\
C1 (fixed stragglers)          & \TODOhpcSH{FA mean$\pm$SD} & \TODOhpcSH{FP mean$\pm$SD} & \TODOhpcSH{$\Delta$} & \TODOhpcSH{p} & \TODOhpcSH{p} \\
C2 (random stragglers)         & \TODOhpcSH{FA mean$\pm$SD} & \TODOhpcSH{FP mean$\pm$SD} & \TODOhpcSH{$\Delta$} & \TODOhpcSH{p} & \TODOhpcSH{p} \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Straggler-tolerance ratio}

To quantify how much each algorithm's performance degrades under system
heterogeneity relative to the no-system-het baseline, we define the
\emph{straggler-tolerance ratio} for algorithm $a$ in condition $c$:
\[
\rho_a^c = \frac{M_a^c}{M_a^{C0}},
\]
where $M$ is mean test macro-F1. Values close to 1 indicate the algorithm
preserves its no-system-het performance; lower values indicate worse
tolerance to stragglers.

\begin{table}[h]
\centering
\caption{Straggler-tolerance ratios. Higher = more tolerant. The prediction
$\rho_\text{FedProx}^c > \rho_\text{FedAvg}^c$ supports the original
FedProx motivation.}
\label{tab:sh-tolerance}
\small
\begin{tabular}{lcc}
\toprule
Condition & $\rho_\text{FedAvg}$ & $\rho_\text{FedProx}$ \\
\midrule
C0 (baseline) & $1.000$ & $1.000$ \\
C1 (fixed stragglers)  & \TODOhpcSH{$\rho_{FA}^{C1}$} & \TODOhpcSH{$\rho_{FP}^{C1}$} \\
C2 (random stragglers) & \TODOhpcSH{$\rho_{FA}^{C2}$} & \TODOhpcSH{$\rho_{FP}^{C2}$} \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Per-class effects under system heterogeneity}

\TODOhpcSH{This subsection will report per-class test F1 for each condition,
using the same Holm-corrected paired Wilcoxon protocol as the statistical-
heterogeneity section. The expectation is that the minority classes most
sensitive to client drift (melanoma in particular) will be most helped by
FedProx under system heterogeneity, paralleling but exceeding the
statistical-heterogeneity per-class result ($\Delta_\text{melanoma} = +0.114$,
$p_{\text{Holm}} = 0.041$).}

%==============================================================================
\section{Discussion}
%==============================================================================

\TODOhpcSH{This section will discuss whether the H1 and H2 results match
the pre-registered predictions:

\begin{itemize}
\item If H1 confirmed: FedProx's advantage under non-IID data extends to
the system-heterogeneity regime; the proximal regulariser is broadly
useful for federated medical-imaging deployment.
\item If H2 confirmed: the system-heterogeneity advantage is larger than
the statistical-heterogeneity advantage, consistent with FedProx's
original theoretical motivation in \citet{Li et al., 2020} that the
proximal term was introduced primarily to handle inexact local updates
from stragglers.
\item If H2 not confirmed but H1 confirmed: FedProx remains useful under
system heterogeneity but the magnitude is similar to the
statistical-heterogeneity regime --- the two effects do not combine
multiplicatively, suggesting the proximal term's primary contribution is
constraining drift regardless of its source.
\item If H1 not confirmed: a substantive negative finding worth
reporting; the statistical-heterogeneity claim is unaffected but the
canonical system-heterogeneity FedProx claim does not transfer to medical
imaging in our setup.
\end{itemize}}

%==============================================================================
\section{Summary of Heterogeneity Experiments}
%==============================================================================

This section closes a two-experiment investigation:

\begin{table}[h]
\centering
\caption{Summary of the two heterogeneity experiments. Each tests a
distinct claim about FedProx and uses the same statistical machinery
(paired Wilcoxon, 10 seeds, balanced\_paired\_7\_clients partition).}
\label{tab:sh-summary}
\small
\begin{tabular}{lll}
\toprule
& Statistical heterogeneity (preceding section) & System heterogeneity (this section) \\
\midrule
What varies & Partition (class allocation) & Per-client local-epoch budget \\
Tests FedProx claim & Drift mitigation under non-IID & $\gamma$-inexact updates from stragglers \\
Headline result & $\Delta = +0.027$, $p = 0.020$ & \TODOhpcSH{pending} \\
Effect size & $r_{rb} = +0.818$ & \TODOhpcSH{pending} \\
\bottomrule
\end{tabular}
\end{table}

%==============================================================================
\section*{References}
%==============================================================================
\begin{description}\setlength{\itemsep}{2pt}
\item Li, T., Sahu, A.~K., Zaheer, M., Sanjabi, M., Talwalkar, A., \&
Smith, V. (2020). Federated optimization in heterogeneous networks. In
\textit{Proceedings of Machine Learning and Systems}, \textit{2},
429--450.
\item McMahan, B., Moore, E., Ramage, D., Hampson, S., \& y~Arcas,
B.~A. (2017). Communication-efficient learning of deep networks from
decentralized data. In \textit{Proceedings of AISTATS} (pp.\ 1273--1282).
\item Wang, J., Liu, Q., Liang, H., Joshi, G., \& Poor, H.~V. (2020).
Tackling the objective inconsistency problem in heterogeneous federated
optimization. \textit{NeurIPS}, 33, 7611--7623.
\item Karimireddy, S.~P., Kale, S., Mohri, M., Reddi, S.~J., Stich, S.~U.,
\& Suresh, A.~T. (2020). SCAFFOLD: Stochastic controlled averaging for
federated learning. In \textit{ICML} (pp.\ 5132--5143).
\item Marija, N. (2025). Federated Learning and Class Imbalances: Breast
Lesion Segmentation in DCE-MRI. MPhil dissertation, University of
Cambridge.
\end{description}

\end{document}
